---
title: "Data visualization"
author: "Elise Gonzales"
date: "2022-08-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(janitor)

#wrappers around ggplot
library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)
library(gapminder)
library(sf)
```

###Read in lizard data    
```{r}
lizards <- read_csv(here("data_tidy", "lizards.csv"))
```

###Basic ggplot - point
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point()
```

##Histogram
```{r}
ggplot(data = lizards, aes(x = total_length)) +
  geom_histogram(color = "purple",
                 fill = "orange",
                 size = 1,
                 linetype = "dotted") #sturgis rule? to figure out reasonable # of bins to accurately represent distribution of data
```

```{r}
ggplot(data = lizards,
       aes(x = total_length, y = weight)) +
         geom_point(
           shape = 24,
           color = "cyan4",
           fill = "chartreuse")
```

##map based on variables
```{r}
ggplot(data = lizards,
       aes(x = total_length, 
           y = weight)) +
  geom_point(aes(color = common_name, 
                 size = total_length), alpha = 0.5) + theme_minimal()
```

##facet
```{r}
ggplot(data = lizards,
       aes(x = total_length, 
           y = weight)) + 
  geom_point() +
  facet_wrap(~common_name)
```

###Re-order
```{r}
lizard_counts <- lizards |> 
  group_by(common_name) |> 
  summarize(counts = n())

#same thing
lizards_total <- lizards |> 
  count(common_name, site, tail) #you can add these extra variables in group_by() too
```

##Column/bar graph, re-order
```{r}
ggplot(data = lizard_counts, 
       aes(y = fct_reorder(common_name, counts), x = counts)) + geom_col()

#convert to factor to specify which order to display variables in, instead of default alphabetical

```

Create scatterplots, faceted by species, for total length vs weight. Color & shape of points should differ based on common_name. Change theme. Make each faceted panel have flexible axes limits. Scale should be free
```{r}
ggplot(data = lizards,
       aes(x = total_length, 
           y = weight)) + 
  geom_point(aes(color = common_name, 
                 shape = common_name,
             alpha = 0.6,
             size = 1)) +
  facet_wrap(~common_name, scales = "free") +
  theme_linedraw() +
  labs(x = "Total length (mm)",
       y = "Weight (g)") +
  title = "Jornada Basin lizard sizes" +
  subtitle = "All lizards follow the standard length-weight relationship, with Western Whiptails being largest" +
  caption = "Data: Jornada Basin LTER, CCBY"
```

#Maps - scales
```{r}
ggplot(data = lizards,
       aes(x = total_length, 
           y = weight)) + 
  geom_point(aes(color = weight)) +
  scale_color_gradientn(colors = c("purple2", "cyan4", "chartreuse", "yellow", "orange"))
```
###Diverging color scheme
```{r}
ggplot(data = lizards,
       aes(x = total_length, 
           y = weight)) + 
  geom_point(aes(color = total_length)) +
  scale_color_steps2(low = "purple", mid = "white", high = "orange", midpoint = 150, breaks = c(50, 75, 150, 180, 200, 220, 280))
```
###Sequential color scheme
```{r}
ggplot(data = lizards,
       aes(x = total_length, 
           y = weight)) +
  geom_point(aes(color = total_length)) +
  scale_color_stepsn(colors = c("yellow", "red", "purple"), breaks = seq(from = 0, to = 300, by = 60))
```
Make a horizontal boxplot with common name on y_axis, total length on x-axis, with boxplot fill changing based on common name. 
```{r}
#create a new column, overwrite existing col. Makes common_name a factor. Puts values in order based on median of total length for each level of species by common name. 
lizards_median <- lizards |> 
  mutate(common_name = fct_reorder(common_name, total_length, .fun = median))

ggplot(data = lizards_median,
       aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name))  

ggplot(data = lizards_median, aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name),
  show.legend = FALSE) +
  scale_fill_paletteer_d(palette = "nbapalettes::suns_city") +
  theme_minimal()
  
```

#other customizations
```{r}
ggplot(data = lizards,
       aes(x = total_length, 
           y = weight)) +
  geom_point(aes(color = weight)) +
  theme(panel.grid.major.y = element_line(color = "pink"),
        panel.grid.minor.y = element_blank(),
        axis.title.x = element_text(color = "orange"),
        axis.title.y = element_text(color = "purple")) +
  annotate("text", x = 300, y = 50, label = "YEW", color = "grey") +
  geom_vline(xintercept = 250, linetype = "dashed", color = "orange", size = 4)
  
  
    
```

```{r}
wws_lizards <- lizards |> 
  filter(common_name == "western whiptail",
         site == "sand")

ggplot(data = wws_lizards,
       aes(x = total_length, 
           y = weight)) +
  geom_point() +
  geom_text_repel(aes(label = toe_num))

gapminder |> 
  filter(year == 2002, continent == "Europe") |> 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  

ggplot(data = lizards,
       aes(x = total_length, 
           y = weight)) +
  geom_point() +  
  gghighlight(toe_num == 250, label_key = toe_num)
  
```

##Map
```{r}
#sf = simple features. spatial data
jornada_veg <- read_sf(here("data_raw", "spatial_vegetation", "doc.kml"))

#Make a map!
ggplot(data = jornada_veg) + 
  geom_sf(data = jornada_veg,
          aes(fill = Name),
          color = NA) +
  theme_minimal() +
  scale_fill_paletteer_d(palette = "ggthemes::manyeys")
```

#Heat-map:tiled plot
```{r}
lizard_counts <- lizards |> 
  mutate(date = lubridate::mdy(date)) |> 
  count(year = lubridate::year(date), common_name) |> 
  drop_na()
```

```{r}
ggplot(data = lizard_counts, aes(x = year, y = common_name)) +
  geom_tile(aes(fill = n), show.legend = FALSE) +
  geom_text(aes(label = n), color = "white", size = 3) +
  scale_fill_gradientn(colors = c("navy", "red", "orange")) +
  theme_minimal()
```

##Beeswarm plot: better than jitter & violin? can only handle so many points...if your sample sizes are huge, not the best option
```{r}
whiptails <- whiptails |> 
  filter(common_name == "western whiptail") |> 
  drop_na(total_length, weight)
```

```{r}
whip_plot1 <- ggplot(data = whiptails, aes(x = sex, y = weight)) + geom_beeswarm() +
  geom_boxplot(fill = NA) 
```

## Marginal plot
```{r}
whip_plot <- ggplot(data = whiptails, aes(x = total_length, y = weight)) + geom_point(aes(color = sex)) + 
  theme(legend.position = "bottom")

ggMarginal(whip_plot, type = "boxplot", groupColour = TRUE)
```

## Patchwork
```{r}
((whip_plot1 + whip_plot) / whip_plot1) + whip_plot & theme_minimal()

```

